Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2023-01-08 edubots_surveys.Rnw

% The Use of Mobile Instant Messaging (MIM) Platforms in Higher Education
The widespread and rapid adoption of free Mobile Instant Messaging
(MIM) tools/platforms such as WhatsApp, Telegram, WeChat and Facebook
Messenger stems from their simplicity, ease of use and multi-modality
(i.e. video, audio, text) \cite{tang2017mobile}. Using these tools in
higher education can facilitate the delivery of personalized learning
that occurs anytime anywhere, promote collaborative learning
experiences and group discussions \cite{panah2020study}.

WhatsApp is, at least in most Western countries, the most popular MIM platform
used by educators to \textcolor{black}{provide} assignments' feedback to students, support course
discussions, and provide learning resources in informal learning settings
\cite{panah2020study}.  Moreover, the use of WhatsApp in higher education could
enhance social presence \cite{tang2017mobile} and foster trust relationships
between educators and students embedded in the social learning process
\cite{gachago2015crossing}; however, this last paper also reflects the need for
learners to "take ownership of the tool" and the advantages of social learning
in general. At the same time, it also mentions different challenges, among which
the most important is the blurring of social and academic life; indeed, there are challenges when using MIM tools that occur due to the
blurring of boundaries between academic and private life. This can lead to
technostress \cite{gachago2015crossing}, difficulty in managing
responsibilities, especially among mature students, and lack of privacy
\cite{tang2017mobile}.  Students' dropout of the MIM groups, as they can leave
groups at any time, can hinder their learning and undermine educators' efforts
\cite{mwakapina2016whatsapp}.  In addition, there is a need to set rules and
norms for these MIM groups in order to maintain the safety of these online
communities for students \cite{abbas2021onlinechat}. However, these rules should
not affect students' ownership and control, since it is vital to advance in
their learning \cite{gachago2015crossing} process.  This is why examining the
role of MIM in higher education is still a challenge, and why the opinions of
the teaching community towards them have to be examined, as we do in this paper.

% The Use of Chatbots in Higher Education
The use of MIM, although possibly valuable by itself, can be enhanced via the
use of chatbots, which, being conversational agents, usually dwell in systems
where synchronous conversations take place.  The use of conversational agents
(chatbots) in higher education is still at its infancy
\cite{yang2019opportunities}. Nevertheless, recent studies examining their
positive impact on students' academic performance \cite{perez2020rediscovering}
and engagement \cite{differ.chat,abbas2021onlinechat} have led to a growing
interest in using this technology in the (possibly virtual) classroom. Indeed,
the use of chatbots in collecting course feedback from students in higher
education improved students' response quality and boosted their enjoyment levels
\cite{abbas2021Surveys}.  According to \cite{Roblyer2010}, using either tools
such as mobile devices or teaching strategies based on gamification
\cite{Yildirim2017} can improve student motivation. In this sense, the authors
in \cite{Pimmer2019} \textcolor{black}{adopted a quasi-experimental, survey-based approach to
report the positive impact of using instant messaging tools in boosting
students' knowledge and mitigating their feelings of isolation}.

Several higher education chatbots' evaluation studies have been undertaken. For
instance, a recent evaluation review study presented by Smutny and Schreiberova
\cite{Smutny2020} examined 47 educational chatbots implemented in Facebook
Messenger with the focus to identify characteristics and quality metrics such as
language, subject matter and platform\textcolor{black}{, whereas the study undertaken by Pérez and collaborators
\cite{perez2020rediscovering} aimed to categorize educational chatbots, according
to their, purpose into service-oriented and teaching-oriented}. The first category
includes those that provide service support such as the chatbot Ask Holly
\cite{meetholly} and Dina \cite{santoso2018dinus}; both chatbots respond to
students' questions about enrollment and registration. Ask L. U.
\cite{LancasterUniversity_2019} answers students' frequently asked questions
about timetables, grades, tutors and societies. LISA \cite{dibitonto2018chatbot}
and Differ \cite{differ.chat} facilitate breaking the ice between new students
by introducing them to each other. Ranoliya et al. \cite{ranoliya2017chatbot}
proposed a generic chatbot for university students that is able to answer to
their frequently asked questions \textcolor{black}{and, for some time, the University of Granada had
\textcolor{black}{Elvira} as a chatbot in its main web page \cite{MOREO20129} to perform that duty.} Besides being able
to answer pre-established (frequent) questions, it did so from the website of
the University of Granada using an inset \textcolor{black}{persona, voiced by a real person, who
lip-synced the answers. The emphasis in the case of Elvira, however, was on the
authenticity of the synchronization more than in making the update of questions
answered simpler, or more interactive, or more open to the rest of the university staff, using actual messaging platforms. In
most cases, the embedded search engine was able to provide more up-to-date
and accurate answers than the ones provided by Elvira, which had to be updated
by hand (and not too often). It was eventually discontinued, and its technology was not adapted to other platforms, since it did not really provide an useful service. Its service was not tied to teaching anyway; it was more related to directory queries and university-wide administrative questions, so even if it is strictly an example of a chatbot in a university setting, it is not really a chatbot that can be adopted by teachers to affect the learning outcomes, something which is the focus of this paper}.

On the other hand, teaching-oriented chatbots are more sophisticated, as they
set personalized learning outcomes and monitor learning progress. For instance,
\cite{fernoagua2018intelligent} reported on ``eduAssistant'', a virtual teaching
assistant chatbot developed on the Telegram messaging platform. In this study,
the Telegram platform is chosen because it is easy to use, students are familiar
with its features, and it enables them to exchange messages in different formats
(text, audio and video) \cite{fernoagua2018intelligent}. In addition, Telegram
could operate on all devices and operating systems. The ``eduAssistant'' chatbot
acts as an automatic agent in teacher-content-student, facilitating real-time
feedback loops and providing a personalized learning experience relevant to the
students' acquired skills and knowledge. Using this chatbot, educators can
create interactive instances in their lectures where they pose questions to
their students and the chatbot assists those who need further help by giving
them more hints and reporting it to their educator's dashboard
\cite{fernoagua2018intelligent}. This can help educators locate those
students that need more attention and send them more educational resources
relevant to their academic attainment. \textcolor{black}{In addition, a recent study
has analysed how the use of chatbots positively affected the learning outcomes
of students in a Chinese class \cite{doi:10.1177/0735633120929622}}.

Despite its (arguably) successful implementation in different higher education
institutions, over all in the fringes of educational activity, not in the actual
student-teacher interaction, its implementation or deployment is not trivial.
Some authors, for instance \cite{sjostrom2018designing} have proposed a
conceptual architecture for the adoption of teaching-oriented chatbots in higher
education.  This conceptual architecture is based on a systematic literature
review of previous studies, examining the design of chatbots in higher
education, as well as making a content analysis of student emails and discussion
forum posts of four instances of a Java programming course.  The study outlined
several design considerations; among them, the authors emphasized the importance
of developing chatbots in platforms that students and educators are familiar
with and can easily access (i.e. Facebook Messenger) which was confirmed with
\cite{hobert2019you} and \cite{fernoagua2018intelligent} studies. In addition, \cite{sjostrom2018designing} argued that a
conceptualization of learners' questions could aid designers in integrating the
appropriate types of questions that the chatbots should support for different
courses.  \textcolor{black}{Some other authors}, like \cite{coronado2018cognitive}, proposed agents that
store learning materials to be provided on demand to students, whereas the
authors in \cite{crockett2017predicting} reported on tutoring systems, which can
perform initial assessments of students' understanding and provide learning
material that would advance their understanding to the next level.

Regarding the factors for the adoption of chatbots in higher education, many
studies have focused on the evaluation of technology acceptance and usability
\cite{Roblyer2010,Pimmer2019}. However, higher education is a special domain
where, according to \cite{hobert2019you}, specific pedagogical factors
such as learning success and increased motivation are more important.  Therefore,
to develop \textcolor{black}{effective chatbots for higher education, all stakeholders'
(i.e. educators, students, institutions, etc) needs should be carefully
collected and taken into consideration
\cite{sjostrom2018designing,tsivitanidouusers}. These needs include, but are not limited to, student's learning success due to higher motivation, but these are a posteriori effects that cannot be assessed in advance; this is why we focus in this paper on what the teachers are looking for in a TAM  (Technology Acceptance Model) and associated technologies like chatbots; both authors focus on what the teachers need in terms of the latter; this paper will focus on a wider perspective trying to ascertain what they are looking for in terms of general messaging technology. The classical literature \cite{tam91} already proposes that any adoption of technology must be tuned in to the user needs and experience. In this paper we will try to find out what those are in order to propose a successful model of adoption of chatbot technology}.

Along this line of research, our previous work \cite{MoraChatbots2021:anon}
aimed at analyzing the expectations of students on this regard; recent papers
have also analysed how the use of chatbots affected the learning outcomes of
students in a Chinese class \cite{doi:10.1177/0735633120929622}. This paper
analysed conversational chatbots in an one-to-one setting, finding that learning
really benefited from it. This is, once again, a proof of the benefit of
chatbots in certain settings; however, there are some prior experiences, as well as needs, that might
prevent the successful acceptance of the technology, and these are what we are
trying to find out in this paper, together with what kind of features would improve
its acceptance.

Besides, the present work is focused on the other key actor in this challenge,
i.e. educators. \textcolor{black}{How they accepted chatbots} has been studied very recently by Chocarro et
al. \cite{doi:10.1080/03055698.2020.1850426}, who, by analyzing surveys, created
a {\em technology acceptance model} that proposed a series of features that
would make chatbots easier to accept; among them, formality of language as well
as easiness and usefulness. The survey targeted primary and secondary education
teachers, and was also more interested in the general use of chatbots in
education, not specifically in a classroom setting, seeking educational
outcomes. However, they establish a series of results that are obviously
interesting and relevant for this work.

We will present next the methodology we have followed to find out teachers'
opinions.

>>>>

% The Use of Mobile Instant Messaging (MIM) Platforms in Higher Education
The widespread and rapid adoption of free Mobile Instant Messaging
(MIM) tools/platforms such as WhatsApp, Telegram, WeChat and Facebook
Messenger stems from their simplicity, ease of use and multi-modality
(i.e. video, audio, text) \cite{tang2017mobile}. Using these tools in
higher education facilitates the delivery of personalized learning
that occurs anytime anywhere, and promotes collaborative learning
experiences and group discussions \cite{panah2020study}.

WhatsApp is, at least in most Western countries, the most popular MIM platform
used by educators to provide assignments' feedback to students, support course
discussions, and contribute resources in informal learning settings
\cite{panah2020study}.  Moreover, the use of WhatsApp in higher education could
enhance social presence \cite{tang2017mobile} and foster trust relationships
between educators and students embedded in the social learning process
\cite{gachago2015crossing}. However, this last paper also reflects the need for
learners to "take ownership of the tool" and the advantages of social learning
in general. At the same time, it also mentions different challenges, among which
the most important is the blurring of social and academic life; indeed, there are challenges when using MIM tools that occur due to the
blurring of boundaries between academic and private life. This can lead to
technostress \cite{gachago2015crossing}, difficulty in managing
responsibilities, especially among mature students, and lack of privacy
\cite{tang2017mobile}.  Students' dropout of the MIM groups, as they can leave
groups at any time, can hinder their learning and undermine educators' efforts
\cite{mwakapina2016whatsapp}.  In addition, there is a need to set rules and
norms for these MIM groups in order to maintain the safety of these online
communities for students \cite{abbas2021onlinechat}. However, these rules should
not affect students' ownership and control, since it is vital to advance in
their learning \cite{gachago2015crossing} process.  This is why examining the
role of MIM platforms in higher education is still a challenge, and why the opinions of
the teaching community towards them have to be examined, as we do in this paper.

% The Use of Chatbots in Higher Education
The use of MIM platforms, although possibly valuable by itself, can be enhanced with the
use of chatbots, which are conversational agents that normally dwell 
synchronous conversations systems. The use of chatbots in higher education is still in its early stages
\cite{yang2019opportunities}. Nevertheless, recent studies have shown their
positive impact on students' academic performance \cite{perez2020rediscovering}
and engagement \cite{differ.chat,abbas2021onlinechat}, leading to a growing
interest in using this technology in the (possibly virtual) classroom. Indeed,
the using chatbots to collect course feedback from students in higher
education improved response quality and boosted enjoyment levels
\cite{abbas2021Surveys}.  According to \cite{Roblyer2010}, the use of
mobile devices or teaching strategies based on gamification
\cite{Yildirim2017} can improve student motivation. In this sense, the authors
in \cite{Pimmer2019} adopted a quasi-experimental, survey-based approach to
report the positive impact of using instant messaging tools in boosting
students' knowledge and mitigating their feelings of isolation.

Several studies have been conducted to evaluate chatbots in higher education. For
instance, a recent study was presented by Smutny and Schreiberova
\cite{Smutny2020} reviewing 47 educational chatbots implemented in Facebook
Messenger, with a focus on identifying characteristics and quality metrics such as
language, subject matter, and platform. Pérez and collaborators
\cite{perez2020rediscovering} attempted to categorize educational chatbots attending to their purpose 
as service-oriented or teaching-oriented. Service-oriented chatbots
include those that provide service support, such as Ask Holly
\cite{meetholly} and Dina \cite{santoso2018dinus}, which responds to
students' questions about enrollment and registration. Ask L. U.
\cite{LancasterUniversity_2019} answers students' frequently asked questions
about schedules, grades, tutors, and societies. LISA \cite{dibitonto2018chatbot}
and Differ \cite{differ.chat} facilitate breaking the ice between new students
by introducing them to each other. Ranoliya et al. \cite{ranoliya2017chatbot}
proposed a generic chatbot for university students that is able to answer 
frequently asked questions. The University of Granada had
Elvira, a built-in chatbot in its main web page \cite{MOREO20129} to perform the same task. In addition to being able
to answer pre-established frequent questions, it did so from the website of
the University of Granada using an inset persona, voiced by a real person, who
lip-synced the answers. However, in the case of Elvira, the emphasis was on the
authenticity of the synchronization rather than making the updated 
of answered questions simpler, more interactive, or more open to the rest of the university staff using actual messaging platforms. In
most cases, the embedded search engine was able to provide more up-to-date
and accurate answers than those provided by Elvira, which had to be updated
by hand (and not too often). It was eventually discontinued, and its technology was not adapted to other platforms, since it did not really provide a useful service. Its service was not tied to teaching anyway; it was more related to directory queries and university-wide administrative questions, so even if it is strictly an example of a chatbot in a university setting, it is not really a chatbot that can be adopted by teachers to affect the learning outcomes, which is the focus of this paper.

On the other hand, teaching-oriented chatbots are more sophisticated, as they
set personalized learning outcomes and monitor learning progress. For instance,
\cite{fernoagua2018intelligent} reported on ``eduAssistant'', a virtual teaching
assistant chatbot developed on the Telegram messaging platform. In this study,
the Telegram platform was chosen because it is easy to use, students are familiar
with its features, and it enables them to exchange messages in different formats
(text, audio and video) \cite{fernoagua2018intelligent}. In addition, Telegram
can operate on all devices and operating systems. The ``eduAssistant'' chatbot
acts as an automatic agent in the teacher-content-student relationship, providing real-time
feedback loops and a personalized learning experience relevant to the
students' acquired skills and knowledge. Using this chatbot, educators can
create interactive instances in their lectures where they pose questions to
their students and the chatbot assists those who need further help by giving
them more hints and reporting it to the educator's dashboard
\cite{fernoagua2018intelligent}. This can help educators identify
students who need more attention and provide them with more educational resources
relevant to their academic achievement. %In addition, a recent study
%has analysed how the use of chatbots positively affected the learning outcomes
%of students in a Chinese class \cite{doi:10.1177/0735633120929622}.

Despite its (arguably) successful implementation in different higher education
institutions, particularly on the fringes of educational activity rather than actual
student-teacher interaction, its implementation or deployment is not trivial.
Some authors, for instance \cite{sjostrom2018designing}, have proposed a
conceptual architecture for the adoption of teaching-oriented chatbots in higher
education.  This conceptual architecture is based on a systematic literature
review of previous studies examining the design of chatbots in higher
education, as well as making a content analysis of student emails and discussion
forum posts from four instances of a Java programming course.  The study outlined
several design considerations; among them, the authors emphasized the importance
of developing chatbots in platforms that students and educators are familiar
with and can easily access (i.e. Facebook Messenger), which was confirmed by
\cite{hobert2019you} and \cite{fernoagua2018intelligent}. In addition, \cite{sjostrom2018designing} argued that 
conceptualizing learners' questions could help designers integrate the
appropriate types of questions that chatbots should support for different
courses. Other authors, such as \cite{coronado2018cognitive}, have proposed agents that
store learning materials to be provided on demand to students, while \cite{crockett2017predicting} 
reported on tutoring systems that can
perform initial assessments of students' understanding and provide learning
materials to advance their understanding to the next level.

Regarding the factors for the adoption of chatbots in higher education, many
studies have focused on the evaluation of technology acceptance and usability
\cite{Roblyer2010,Pimmer2019}. However, higher education is a special domain
where, according to \cite{hobert2019you}, specific pedagogical factors
such as learning success and increased motivation are more important.  Therefore,
to develop effective chatbots for higher education, the needs of all stakeholders
(i.e. educators, students, institutions, etc) should be carefully
collected and taken into consideration
\cite{sjostrom2018designing,tsivitanidouusers}. These needs include, but are not limited to, student learning success due to higher motivation, but these are a posteriori effects that cannot be assessed in advance. This is why we focus in this paper on what  teachers are looking for in a Technology Acceptance Model (TAM) and associated technologies like chatbots. Both authors focus on what teachers need in terms of the latter; this paper will focus on a wider perspective, trying to determine what they are looking for in terms of general messaging technology. The classical literature \cite{tam91} already proposes that any adoption of technology must be tuned to the user needs and experience. In this paper, we will try to find out what those are in order to propose a successful model of adoption of chatbot technology.

In line with this research, our previous work \cite{MoraChatbots2021:anon}
aimed at analyzing the expectations of students in this regard. Recent papers
have also analysed how the use of chatbots affected the learning outcomes of
students in a Chinese class \cite{doi:10.1177/0735633120929622}. This paper
analysed conversational chatbots in a one-to-one setting, finding that learning
truly benefited from it. This is, once again, an evidence of the benefit of
chatbots in certain settings; however, there are some prior experiences, as well as needs, that may
prevent the successful acceptance of the technology, and these are what we are
trying to find out in this paper, alng with what kind of features would improve
its acceptance.

Besides, this work is focused on the other key actor in this challenge: 
educators. How they accept chatbots has been studied very recently by Chocarro et
al. \cite{doi:10.1080/03055698.2020.1850426}, who, by analyzing surveys, created
a TAM that proposed a series of features that
would make chatbots easier to accept, including formality of language as well
as easiness and usefulness. The survey targeted primary and secondary education
teachers and was also more interested in the general use of chatbots in
education, not specifically in a classroom setting, seeking educational
outcomes. However, their results are obviously
interesting and relevant for this work.

We will present next the methodology we followed to find out teachers' opinions.
